---
title: "EDA"
format: html
editor: visual
---

```{r, include=FALSE}
#Clearing the working environment
rm(list=ls())
```

Before beginning the EDA, it is important to load the necessary libraries!

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
```

## Introduction

In this project, I analyze data on diabetes. These data are a subset of the Behavioral Risk Factor Surveillance System (BRFSS) which is an annual telephone survey conducted by the Centers for Disease Control (CDC). This subset includes responses from 253,680 people and was collected in 2015. This dataset contains information on whether the person does not have diabetes or if they have either pre-diabetes or diabetes, in addition to other numeric and categorical health indicators. There are binary, indicator variables describing whether or not the person has a history of high blood pressure, high cholesterol, smoking, stroke, coronary heart disease (CHD) or myocardial infarction (MI), physical activity in the past 30 days, daily fruit consumption, daily veggie consumption, alcohol consumption, healthcare coverage, missing medical care due to cost, and difficulty walking. There are also categorical variables with multiple levels describing the person's general health, mental health, and physical health as well as demographic characteristics like sex, age, education, and income. There is also a numeric variable, BMI or body mass index, calculated by (weight in kg)/(height in meters)\^2.

The goal of this project is to develop a good model to model and make predictions about whether or not a person has diabetes. According to a CDC article published in 2024 (https://www.cdc.gov/diabetes/risk-factors/index.html), being overweight or obese, being 45 or older, having a close relative with Type 2 diabetes, being physically active 3 or fewer times in a week, having non-alcoholic fatty liver disease (NAFLD), having gestational diabetes, or being an African American, Hispanic or Latino, American Indian, or Alaska Native person are main factors associated with prediabetes and Type 2 diabetes. In our dataset, we have access to BMI (overweight or obesity), physical activity, and age. I will explore the summary and distribution of those three variables (BMI, physical activity, and age).

## Data

### Reading in the data

First, we read in the data. The data has been downloaded and stored in my GitHub repository, but was originally sourced from Kaggle (https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data).

```{r}
#Reading in the data
diabetes_data <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")
```

### Cleaning the data

Next, we clean and manipulate the data. We select the four variables we are most interested in which are the response variable (whether or not the person has been diagnosed with pre-diabetes or diabetes) and three explanatory variables. The explanatory variables are whether the person is physically active (a binary variable), the person's BMI (a numeric variable), and the person's age (a categorical variable with 13 levels corresponding to different age groups).

```{r}
#Selecting response and 3 chosen explanatory variables
#Converting categorical variables from numeric variables to factor variables
diabetes_select <- diabetes_data |>
  select(Diabetes_binary, BMI, PhysActivity, Age) |>
  rename("diabetes_resp" = Diabetes_binary) |>
  rename("bmi" = BMI) |>
  rename("phys_activity" = PhysActivity) |>
  rename("age" = Age) |>
  mutate(diabetes_resp = factor(diabetes_resp, levels = c(0, 1), 
                                labels = c("No diabetes", "Prediabetes or diabetes"))) |>
  mutate(phys_activity = factor(phys_activity, levels = c(0, 1), 
                                    labels = c("No physical activity", "Physical activity"))) |>
  mutate(age = factor(age, levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13), 
                      labels = c("Age 18-24", "Age 25-29", "Age 30-34", "Age 35-39", 
                                 "Age 40-44", "Age 45-49", "Age 50-54", "Age 55-59",
                                 "Age 60-64", "Age 65-69", "Age 70-74", "Age 75-79", 
                                 "Age 80 or older")))
```

### Checking for Missingness

```{r}
colSums(is.na(diabetes_select))
```

After selecting our four variables, we check for missingness. None of our 4 variables have any missing values (hooray!) so we can proceed to further data visualizations and summaries.

## Summarizations

### Univariate Explorations

Before looking at the bivariate relationship between the 3 explanatory variables and the response variable, whether or not the person has diabetes, we will look at each of the variables individually.

#### BMI

First, we will look at BMI. BMI is a numeric variable that ranges from 12 to 98 and is calculated by (weight in kg)/(height in meters)\^2.A summary of important statistics for BMI is below.

```{r}
diabetes_select |>
  summarize(across(bmi, .fns = list("mean" = mean,
                                    "median" = median, 
                                    "var" = var,
                                    "sd" = sd,
                                    "IQR" = IQR), .names = "{.col} {.fn}"))
```

In general, the mean and median for BMI are very similar. This indicates that the data are roughly distributed symmetrically rather than skewed. However, we also see that there is decent variance. I am curious to explore that variance further and understand how the variance could potentially relate to diabetes diagnoses.

#### Physical Activity

Second, we will look at the variable for physical activity. This is a binary variable where the value of 1 indicates that the person had done some physical activity outside of their job in the last 30 days while a value of 0 indicates that they had not.

```{r}
addmargins(table(diabetes_select$phys_activity))
```

```{r}
# Creating a Pie Chart to better visualize physical activity distribution 
# Used pie chart instructions from: https://r-graph-gallery.com/piechart-ggplot2.html

#Preparing the data
phys_act_data <- diabetes_select |>
  count(phys_activity) |>
  mutate(percent = n / sum(n) * 100)

# Plot
ggplot(phys_act_data, aes(x = "", y = percent, fill = phys_activity)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Outcome Distribution", x = NULL, y = NULL, fill = "Last 30 Days Physical Activity") +
  theme_void()
```

In general, this table and plot show that most people in the dataset are physically active (it looks like around 75% are). However, 25% are not. I am curious what the distribution of diabetes or prediabetes diagnosis looks like within the two groups.

#### Age Groups

Third, we will look at the last explanatory variable--age groups. This is a categorical variable with 13 levels each reflecting an age group. The 13 groups are: Age 18-24, Age 25-29, Age 30-34, Age 35-39, Age 40-44, Age 45-49, Age 50-54, Age 55-59, Age 60-64, Age 65-69, Age 70-74, Age 75-79, and Age 80 or older.

```{r}
addmargins(table(diabetes_select$age))
```

```{r}
age_groups <- diabetes_select |>
  group_by(age) |>
  summarize(count = n())

ggplot(data = age_groups, aes(x = age, y = count)) +
  geom_bar(stat = "identity") +
  labs(title = "Age Groups", x = "Age Group", 
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This plot shows that there is a gentle left skew in the age group data. Overall, it appears like we have a lot of observations from many different age groups as each group n is at least 5,700. The CDC article stated that diabetes is more common in people aged 45 and above. I am curious how the distribution of diabetes overlaps with the distribution of age groups.

#### Diabetes

Finally, we will look at the response variable. This variable describes whether or not the person has diabetes or pre-diabetes. A value of 1 means yes and a value of 0 means no.

```{r}
addmargins(table(diabetes_select$diabetes_resp))
```

This table shows that not being diagnosed with diabetes is significantly more common than having a pre-diabetes or diabetes diagnosis. I am mostly curious how the distribution of diabetes changes based on the level of predictor variables. I will explore that further in the following contingency tables and bar charts.

### Contingency Tables and Bar Charts

We will use contingency tables and bar charts to assess if there is a difference in diabetes rates at different levels of categorical variables.

#### Physical Activity

First, we will look at diabetes rates among people who do physical activity vs those who do not.

```{r, message=FALSE}
diabetes_select |>
  group_by(diabetes_resp, phys_activity) |>
  summarize(count = n()) |>
  mutate(percentage = count / sum(count) * 100) |>
  pivot_wider(names_from = phys_activity,
              values_from = c(count, percentage),
              names_glue = "{phys_activity} {.value}")
```

```{r}
# Stacked barplot with multiple groups
phys_act_groups <- diabetes_select |>
  group_by(phys_activity, 
           diabetes = factor(diabetes_resp, 
                             levels = c("No diabetes", "Prediabetes or diabetes"))) |>
  summarize(count = n(), .groups = "drop")

ggplot(data = phys_act_groups, 
       aes(x = phys_activity, y = count, fill = diabetes)) +
  geom_bar(stat="identity") +
  labs(title = "Diabetes and Physical Activity", x = "Physical Activity", 
       y = "Count", fill = "Diabetes Response")
```

The counts and percentages displayed in the table and the visual bar chart reveal that diabetes diagnoses are much more prevalent among people who have not exercised outside of their job in the past 30 days.

#### Age Groups

Next, we will look at diabetes rates across the 13 different age groups.

```{r, message=FALSE}
diabetes_select |>
  group_by(diabetes_resp, age) |>
  summarize(count = n()) |>
  mutate(percentage = count / sum(count) * 100) |>
  pivot_wider(names_from = age,
              values_from = c(count, percentage),
              names_glue = "{age} {.value}")
```

```{r}
# Stacked barplot with multiple groups
age_groups <- diabetes_select |>
  group_by(age, 
           diabetes = factor(diabetes_resp, levels = c("No diabetes", "Prediabetes or diabetes"))) |>
  summarize(count = n(), .groups = "drop")

ggplot(data = age_groups, aes(x = age, y = count, fill = diabetes)) +
  geom_bar(stat="identity") +
  labs(title = "Diabetes and Age Groups", x = "Age Group", 
       y = "Count", fill = "Diabetes Response") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The counts and percentages displayed in the table and the visual bar chart reveal that diabetes diagnoses are slightly more prevalent among older people. This effect appears to peak around the ages of 60-80.

### Summarizing Center and Spread

#### BMI

```{r}
diabetes_select |>
  group_by(diabetes_resp) |>
  summarize(across(bmi, .fns = list("mean" = mean,
                                    "median" = median, 
                                    "var" = var,
                                    "sd" = sd,
                                    "IQR" = IQR), .names = "{.col} {.fn}"))
```

```{r}
# Use semi-transparent fill
p <- ggplot(diabetes_select, aes(x = bmi, fill = diabetes_resp)) +
  geom_density(alpha = 0.3) +
  labs(title = "BMI Distribution by Diabetes Diagnosis", 
       x = "BMI", y = "Density", fill = "Diabetes Diagnosis")
p
```

The table of mean BMIs and the plot show that those with pre-diabetes or diabetes have higher BMIs than those who do not.

[Click here for the Modeling Page](Modeling.html)
